Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-26 ind_files/04_field_eqns.tex
Add technical comments on hypotheses needed for existence of density function

\begin{example}
Consider the birth and death functionals:
\begin{equation}
\label{Rogers_logistic_BD}
\begin{aligned}
b(x|\nu) &= r\int\limits_{\mathcal{T}}m(x,y)\nu(y)dy; \ m(x,y) = \exp\left(\frac{-(x-y)^2}{\sigma_{m}^{2}}\right)\\
    d(x|\nu) &= \frac{\nu(x)}{Kn(x)}\int\limits_{\mathcal{T}}\alpha(x,y)\nu(y)dy; \ \alpha(x,y) = \exp\left(\frac{-(x-y)^2}{\sigma_{\alpha}^{2}}\right)
\end{aligned}
\end{equation}
This choice corresponds to an asexual population having a constant (per-capita) birth rate $r$. Birth is sometimes with mutation, and the extent of the mutations is controlled by a Gaussian kernel $m(x,y)$. The death rate is density-dependent, mediated by a Gaussian competition kernel $\alpha(x,y)$, and also contains a phenotype-dependent carrying capacity controlled by $n(x)$, scaled by a constant $K$. The biological interpretation of the death rate is through ecological specialization for limiting resources - Individuals have different intrinsic advantages (controlled by $n(x)$), and experience greater competition from conspecifics that are closer to them in phenotype space (controlled by $\alpha(x,y)$).
\end{example}
Let us now define, for each $x \in \mathcal{T}$, two \emph{step operators} $\mathcal{E}_{x}^{\pm}$ that satisfy
\begin{equation*}
    \mathcal{E}_{x}^{\pm}[f(y,\nu)] =  f(y,\nu \pm \delta_x)
\end{equation*}
In other words, the step operators $\mathcal{E}_{x}^{\pm}$ simply describe the effect of adding or removing a single individual with trait value $x$ from the population.
We will assume that our process is nice enough that the hypotheses of the Radon-Nikodym theorem are satisfied and we can find a density function $P(\nu,t)$ such that the probability that the process takes value $\nu^{(t)}$ at time $t$ is given by $\int_{\mathcal{T}}P(\nu,t)dx$. In one dimension, this is not necessarily a difficult requirement for biologically reasonable choices of the birth and death functionals~(\cite{dawson_stochastic_1975,walsh_introduction_1986,konno_stochastic_1988,reimers_one_1989,dawson_stochastic_2000}; Also see proposition 3.1 in~\cite{champagnat_individual_2008}).

>>>>

\begin{example}
Consider the birth and death functionals:
\begin{equation}
\label{Rogers_logistic_BD}
\begin{aligned}
b(x|\nu) &= r\int\limits_{\mathcal{T}}m(x,y)\nu(y)dy; \ m(x,y) = \exp\left(\frac{-(x-y)^2}{\sigma_{m}^{2}}\right)\\
    d(x|\nu) &= \frac{\nu(x)}{Kn(x)}\int\limits_{\mathcal{T}}\alpha(x,y)\nu(y)dy; \ \alpha(x,y) = \exp\left(\frac{-(x-y)^2}{\sigma_{\alpha}^{2}}\right)
\end{aligned}
\end{equation}
This choice corresponds to an asexual population having a constant (per-capita) birth rate $r$. Birth is sometimes with mutation, and the extent of the mutations is controlled by a Gaussian kernel $m(x,y)$. The death rate is density-dependent, mediated by a Gaussian competition kernel $\alpha(x,y)$, and also contains a phenotype-dependent carrying capacity controlled by $n(x)$, scaled by a constant $K$. The biological interpretation of the death rate is through ecological specialization for limiting resources - Individuals have different intrinsic advantages (controlled by $n(x)$), and experience greater competition from conspecifics that are closer to them in phenotype space (controlled by $\alpha(x,y)$).
\end{example}
Let us now define, for each $x \in \mathcal{T}$, two \emph{step operators} $\mathcal{E}_{x}^{\pm}$ that satisfy
\begin{equation*}
    \mathcal{E}_{x}^{\pm}[f(y,\nu)] =  f(y,\nu \pm \delta_x)
\end{equation*}
In other words, the step operators $\mathcal{E}_{x}^{\pm}$ simply describe the effect of adding or removing a single individual with trait value $x$ from the population.
We will assume that our process is nice enough that we can find a function $P(\nu,t)$ such that the probability that the process takes value $\nu^{(t)}$ at time $t$ is given by\footnote{If you know some measure theory, note that this amounts to saying the following (I think?). Let $(\mathcal{M}(\mathcal{T}), \mathscr{F}, \{\mathscr{F}_t\}_{t\geq 0}, \mathbb{P})$ be the filtered probability space in which our adapted stochastic process $\{X_t\}_{t \geq 0}$ lives. Let $m$ be the Lebesgue measure on $(\mathcal{T},\mathcal{B}(\mathcal{T}))$. Then, we require $\mathbb{P}(X_t \ll m \ \forall \ t) = 1$, \emph{i.e.} we need the process to be $[\mathbb{P}]$-a.s. absolutely continuous with respect to the Lebesgue measure for all time. Since each $\nu \in \mathcal{M}(\mathcal{T})$ is a finite measure, it is obviously $\sigma$-finite, and we can now thus use the Radon-Nikodym theorem to find a density function $R_{m}(x,t)$ such that for any $t \geq 0$ and any $B \in \mathcal{B}(\mathcal{T})$, we have $X_t(B) = \int_B R_{m}(x,t) dm \ [\mathbb{P}]\textrm{-a.s.}$. We also require being able to find a second function $Q_{\mathbb{P}}(\nu,t)$ on $\mathscr{F} \times [0,\infty)$ such that we can write $\mathbb{P}(X_t = \nu) = \int_{\mathcal{T}} Q_{\mathbb{P}}(\nu,t)\nu(dx)$. This is again similar to a density function and thus essentially  amounts to an absolute continuity condition, except on the stochastic process itself (and thus the measure $\mathbb{P}$ defined on $\mathscr{F}$) rather than on its state space (i.e. the measures $\nu \in \mathcal{M}(\mathcal{T})$ defined on $\mathcal{T}$). We can now use the $[\mathbb{P}]$-a.s. assured absolute continuity with respect to the Lebesgue measure of any measure $\nu$ that $X_t$ takes on to write $Q_{\mathbb{P}}(\nu,t)\nu(dx) = Q_{\mathbb{P}}(\nu,t)R_{m}(x,t)dm$. We can now define $P(\nu,t)dm = Q_{\mathbb{P}}(\nu,t)R_{m}(x,t)dm$ to write $\mathbb{P}(X_t = \nu) =  \int_{\mathcal{T}}P(\nu,t)dx$ as desired, where we have changed notation from $dm$ to $dx$ as the usual abuse of notation for the Lebesgue measure.} $\int_{\mathcal{T}}P(\nu,t)dx$. In one dimension, this is often automatically satisfied for biologically reasonable choices of the birth and death functionals~(\cite{dawson_stochastic_1975,walsh_introduction_1986,konno_stochastic_1988,reimers_one_1989,dawson_stochastic_2000}; Also see section 2.3 in~\cite{etheridge_introduction_2000} and proposition 3.1 in~\cite{champagnat_individual_2008}).

### 2023-03-26 ind_files/05_unification_discussion.tex
Add technical comments on hypotheses needed for existence of density function

Importantly, the formalism I develop here in terms of functional Fokker-Planck equations likely does \emph{not} carry over to the study of higher dimensional quantitative traits (or populations which vary in two or more one-dimensional quantitative traits) because these processes are routinely badly behaved in higher dimensions: In particular, a probability density $P(\phi,t)$ with respect to the reals frequently does not even exist in higher dimensions if one has any biologically non-trivial features such as interactions between types (\cite{fleming_measure-valued_1979, walsh_introduction_1986,etheridge_introduction_2000}; Also see for example~\cite{evans_measure-valued_1994}), rendering equation \eqref{disc_functional_field_eqns} entirely meaningless. My (admittedly limited) understanding is that such processes are also rather difficult to study in two or more dimensions from the SPDE side for similar reasons~\citep{etheridge_introduction_2000}, where the lack of a density leads to distribution-valued (rather than function-valued) solutions at best and thus demands a considerable amount of advanced functional analysis to establish even `basic' existence/uniqueness results~\citep{walsh_introduction_1986,carmona_stochastic_1999,prevot_concise_2007,liu_stochastic_2015,balan_gentle_2018}. It may well be the case that concrete biologically useful progress in this direction requires new mathematics altogether, a situation increasingly also encountered in other areas of mathematical biology~\citep{vittadello_open_2022}. Just as theoretical physics has done in the past, the pursuit of general mathematical descriptions of biological theory may therefore also inspire new ideas that are of independent interest to `pure' mathematicians~\citep{cohen_mathematics_2004}.

>>>>

Importantly, the formalism I develop here in terms of functional Fokker-Planck equations likely does \emph{not} carry over to the study of higher dimensional quantitative traits (or populations which vary in two or more one-dimensional quantitative traits) because these processes are routinely badly behaved in higher dimensions: In particular, a probability density $P(\phi,t)$ with respect to the reals frequently does not even exist in higher dimensions if one has any biologically non-trivial features such as interactions between types~\citep{fleming_measure-valued_1979, walsh_introduction_1986,konno_stochastic_1988,evans_measure-valued_1994,etheridge_introduction_2000}, rendering equation \eqref{disc_functional_field_eqns} entirely meaningless. My (admittedly limited) understanding is that such processes are also rather difficult to study in two or more dimensions from the SPDE side for similar reasons~\citep{etheridge_introduction_2000}, where the lack of a density leads to distribution-valued (rather than function-valued) solutions at best and thus demands a considerable amount of advanced functional analysis to establish even `basic' existence/uniqueness results~\citep{walsh_introduction_1986,carmona_stochastic_1999,prevot_concise_2007,liu_stochastic_2015,balan_gentle_2018}. It may well be the case that concrete biologically useful progress in this direction requires new mathematics altogether, a situation increasingly also encountered in other areas of mathematical biology~\citep{vittadello_open_2022}. Just as theoretical physics has done in the past, the pursuit of general mathematical descriptions of biological theory may therefore also inspire new ideas that are of independent interest to `pure' mathematicians~\citep{cohen_mathematics_2004}.
